quicR: An R Library for Streamlined Data Handling of Real-Time Quaking Induced Conversion Assays

### 2025-01-15 man/manuscript/manuscript.qmd
Updated manuscript

## Input Formatting
MARS exports data into Excel workbooks. For many of the quicR functions to work together, the workbooks must be formatted correctly. In MARS, select "Excel Report", and a pop-up window will appear (see [@fig-mars]). The following options must be checked for proper output: "Microplate View", "Table view", "Transpose table", and "Add test run information". After clicking "Export report to Excel", an Excel workbook is created such as in [@fig-marsoutput].

![Excel export settings in MARS](images/MARS_settings.png){#fig-mars height=50%}

@@

Excel spreadsheet exported from MARS.  
:::

## Key Metrics and Calculations
quicR has functions for calculating TtT, MPR, and MS. A graphical representation of these can be found in [@fig-metrics]. There is no dedicated function for RAF since it can simply be expressed as the inverse of TtT, and can therefore be calculated separately as in [@tbl-metrics].

@@

MPR is defined as the maximum fluorescence divided by the background fluorescence. Thus, in order to calculate, the raw data must first be normalized against the background. This is done by choosing a cycle for background determination, and then dividing each read by that value. Once this is done, the MPR is taken as the max value.

MS is determined by approximating the maximum of the derivative of the raw data and is typically reported in units of $\Delta$RFU/h (i.e. the change in relative fluorescent units per hour). Originally, this was accomplished by applying a linear regression to a sliding window; however, this proved to be very computationally expensive. Now, the slopes are calculated using differences between two data points within the range of the sliding window. While this slightly reduces the accuracy of the approximation, the improvement in computation time exceeded the loss in accuracy.

![Example graphs highlighting the different calculated metrics. The red curve represents a raw data curve that has been normalized against background. MPR is calculated as the maximum fluorescent value achieved in the normalized raw data. TtT is determined as the time required to cross a given threshold (in this example, the threshold is 0.2). The blue curve represents the derivative of the raw data, and max slope is determined as the maximum of the derivative.](images/metric_example.png){#fig-metrics}

@@

The project was managed using Git for version control, enabling efficient tracking of changes and fostering collaborative development. The source code is hosted on GitHub, ensuring transparency and encouraging contributions from the research community. GitHub Actions were employed to automate testing and deployment workflows, verifying that the package remains compatible with current versions of R and its dependencies.

## Testing and Validation
Robust testing protocols were implemented to guarantee the reliability of quicR’s functionality. Each function was subjected to comprehensive unit testing using the testthat package [@testthat2011]. These tests ensure that key metrics, data manipulation routines, and visualization tools perform as expected, even as the package evolves.

@@

Another avenue for improvement lies in the standardization of RT-QuIC diagnostic criteria. quicR provides tools to calculate key metrics, but consensus on thresholds and interpretations remains a challenge for the field. Collaborative efforts among researchers and clinicians are necessary to define universal criteria, enabling quicR to fully realize its potential as a diagnostic aid.

# Conclusion
quicR offers a powerful solution for the analysis and visualization of RT-QuIC data, addressing critical needs in a rapidly evolving field. By enabling consistent data handling and interpretation, quicR lays the groundwork for improved diagnostic consistency and reproducibility. The package's open-source nature ensures that it will continue to evolve, integrating new insights and technologies as they emerge.

@@

# Acknowledgments

>>>>

## Input Formatting
The FLUOstar® Omega series microplate readers (BMG Labtech, Ortenberg, Germany) are by far the most common readers used for RT-QuIC. As such, their analysis software, MARS, is integral to data input for this package. MARS exports data into Excel workbooks, and for many of the quicR functions to work together, the workbooks must be formatted correctly. In MARS, select "Excel Report", and a pop-up window ([@fig-mars]) will appear. The following options must be checked for proper output: "Microplate View", "Table view", "Transpose table", and "Add test run information".

![Excel export settings in MARS. Ensure that "Microplate View", "Table view", "Transpose table", and "Add test run information" are selected.](images/MARS_settings.png){#fig-mars height=50%}

After clicking "Export report to Excel", an Excel workbook is created such as in [@fig-marsoutput]. This workbook will contain two sheets, the first with the microplate views and the second with the table view. The first sheet will have all of the relevant metadata while the second sheet will have all of the raw fluorescent data.

@@

Excel workbook exported from MARS.  
:::

## Key Metrics and Calculations
quicR has functions for calculating TtT, MPR, and MS. A graphical representation of these can be found in [@fig-metrics]. There is no dedicated function for RAF since it can simply be expressed as the inverse of TtT, and can therefore be calculated separately as in the example in [@tbl-metrics].

@@

MPR is defined as the maximum fluorescence divided by the background fluorescence. Thus, in order to calculate, the raw data must first be normalized against the background. This is done by choosing a cycle for background determination, and then dividing each read by that value. The MPR is taken as the max value of the normalized data.

MS is determined by approximating the maximum of the derivative of the raw data and is typically reported in units of $\Delta$RFU/h (i.e. the change in relative fluorescent units per hour). Originally, this was accomplished by applying a linear regression to a sliding window; however, this proved to be very computationally expensive. Now, slopes are calculated using differences between two data points within the range of a sliding window. While this slightly reduces the accuracy of the approximation, the improvement in computation time exceeded the loss in resolution.

![Example graphs highlighting the different calculated metrics. The red curve represents a raw data curve that has been normalized against background. The level has also been decreased by 1 to better show the relationship with the derivative. MPR is calculated as the maximum fluorescent value achieved in the normalized raw data. TtT is determined as the time required to cross a given threshold (in this example, the threshold is 0.2). The blue curve represents the derivative of the raw data, and max slope is determined as the maximum of the derivative.](images/metric_example.png){#fig-metrics}

@@

The project was managed using Git for version control, enabling efficient tracking of changes and fostering collaborative development. The source code is hosted on GitHub (https://github.com/gage1145/quicR), ensuring transparency and encouraging contributions from the research community. GitHub Actions were employed to automate testing and deployment workflows, verifying that the package remains compatible with current versions of R and its dependencies.

## Testing and Validation
Robust testing protocols were implemented to guarantee the reliability of quicR’s functionality. Each function was subjected to comprehensive unit testing using the testthat package [@testthat2011]. These tests ensure that key metrics, data manipulation routines, and visualization tools perform as expected, even as the package evolves. Each new build of quicR must pass these tests in order to be released.

@@

Another avenue for improvement lies in the standardization of RT-QuIC diagnostic criteria. quicR provides tools to calculate key metrics, but consensus on thresholds and interpretations remains a challenge for the field. Collaborative efforts among researchers and clinicians are necessary to define universal criteria, enabling quicR to fully realize its potential as a diagnostic aid. Diagnostic determinations could easily be built into the library, but a larger consensus within the research community will need to be reached to warrant inclusion.

# Conclusion
quicR offers a powerful solution for the cleaning, analysis, and visualization of RT-QuIC data, addressing critical needs in a rapidly evolving field. By enabling consistent data handling and interpretation, quicR lays the groundwork for improved diagnostic consistency and reproducibility. The package's open-source nature ensures that it will continue to evolve, integrating new insights and technologies as they emerge.

@@

# Acknowledgments
Special thanks to Beni Altmann at The Comprehensive R Archive Network (CRAN) for help during the submission process to CRAN.
